Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-08-02 Manuscript/MPESubmission/Revision/response.tex

"Until now, only few attempts have been made to study study the impact of this missing data issue on phylogenetic inference in a Total Evidence framework (i.e. using combined molecular and morphological data; e.g. Wiens et al., 2005; Pattinson et al., 2014).
For example, Wiens et al. (2005) only assess the effect of missing data by comparing the node support between the complete dataset that contains missing data and sub-datasets with no missing data.
Furthermore, Pattinson et al. (2014) assess the effect of removing both molecular and morphological data (“artificial extinction”) on the topology.
These studies, however, are limited by their empirical-only approach (Wiens et al., 2005) and the scheme of generating missing data (Pattinson et al., 2014)."

"In this study, we propose a theoretical assessment of the effect of missing data in the Total Evidence method by thoroughly eliminating living or fossil data or all data overall.
We test the effect of missing data by measuring two crucial aspects of topology in both Maximum Likelihood and Bayesian Inference methods: (i) the conservation of clades and (ii) the displacement of wild-card taxa."

% What was left open (56)

% Conclusion (52)

Therefore we advise that one should focus on having morphological characters coded for a large amount of living taxa present in the matrix (i.e. 50\%) for accurately combining both living and fossil species in phylogenies.
Doing so grants a good overlap of morphological characters between living and fossil taxa, allowing the fossil taxa to be positioned relatively to the living taxa based on their actual shared derived characters rather than on simply available data."

Additionally, the topology of the Bayesian consensus trees, regardless the amount of missing data, were always closer to the ``best'' tree topology than the Maximum Likelihood trees.
This has also been observed in empirical data (e.g. Arcila et al., 2015) were Maximum Likelihood trees inferred from a Total Evidence matrix were less supported than the Bayesian consensus tree.
This might have an important impact on estimating topologies in the Total Evidence framework since previous studies had to rely either on molecular scaffolds (e.g. Slater, 2013), taxonomic constraints (e.g. Slater, 2013; Beck and Lee, 2014) or even by fixing the topology (e.g. Ronquist et al., 2012a).
Therefore, we suggest extracting such topological frames from the Bayesian consensus tree if needed.
To conclude, the results of our analyses are encouraging and show that it is possible to accurately combine both neontological and palaeontological data in the same phylogeny as long as both types of data sufficiently overlap.
Hopefully, using these approaches will greatly improve our understanding of macroevolutionary patterns and processes."
%Original: Additionally, the topology of the Bayesian consensus trees, regardless the amount of missing data, were always closer to the “best” tree topology than the Maximum Likelihood trees. Therefore, we advise using Bayesian consensus tree topologies if topology must be fixed for furhter analyses. The results of our analyses are encouraging and show that it is possible to combine both neontological and palaeontological data in the same phylogeny despite issues of missing data. Hopefully, using these approaches will greatly improve our understanding of macroevolutionary patterns and processes.

See also reviewer 1 specific comment 52 for details on the choice of citing Arcial et al 2015.

% -----------------
% Comment 12
% -----------------
\item{\textcolor{blue}{\textbf{Figure 4:} Figure 4 has a color scheme issue, the colors in the caption do not match up to the figure.}}

@@

We revised the statements in the whole paragraph as follows:
"Another practical implication of our results regards the tree inference methods.
Because the Bayesian consensus trees consistently recovered topologies closer to the ``best'' tree topology than the Maximum Likelihood trees, we advice using the Bayesian consensus trees as a topological constraint for tree inferences using the Total-Evidence method such as tip-dating (e.g. Ronquist et al., 2012a; Wood et al., 2013; Matzke, 2014; although it is possible that including dating information during tree inference could also improve the accuracy of the Bayesian posterior tree distribution).
Additionally, using the Bayesian consensus tree rather than the Maximum Likelihood can reduce the amount of ``false positive'' topologies.
As shown in figure 5 and discussed in the section above (Effects of tree inference methods), the Bayesian consensus tree is more likely to not resolve nodes weakly supported due to missing data than the Maximum Likelihood tree that is more likely to wrongly resolve such nodes (i.e. creating a ``false positive'' node).
Note, however, that we do not suggest discarding the Bayesian posterior tree distribution once one chose to use the Bayesian consensus tree topology even though they performed poorly in recovering the ``best'' tree topology in our simulations (this can probably be imputed to the difficulties comparing distributions of trees; see above).
This can be particularly important because these trees will be invaluable for phylogenetic comparative analyses. For example a sub-sample of posterior trees distributions can be used to asses macroecological questions while better taking into account topological uncertainty (e.g. Fritz et al. 2009 and Jetz et al. 2012 trees used in Healy et al. 2014)."

% -----------------
% Specific Comment 50
% -----------------
\item{\textcolor{blue}{\textbf{line 612-613:} too obvious and not especially helpful. We should aim to have as little missing data as possible…more helpful, and something especially discovered in this study, is concentrating on having overlapping morphological characters between the living and fossil taxa to be able to place the living taxa with strong support via large datasets (combined data) and being able to position the fossils based on their shared derived characters with living taxa. This is an important point to make clearer. }}

We are thankful for this comment as, indeed, our suggestion was a bit to obvious and not really helpful. See reviewer 1 comment 11 for the changes to the conclusion.

% -----------------
% Specific Comment 51
% -----------------
\item{\textcolor{blue}{\textbf{line 613-615:} rephrase - sentence is choppy with all the commas. }}

@@

Unfortunately, we didn't found many studies formally testing the important questions asked by the reviewer. However, there exist few empirical analysis that contain results supporting our claims.
Firstly, for the question about the topological differences between Maximum Likelihood and the Bayesian consensus tree, in Arcila et al 2015 (MPE)'s supplementary analysis, the authors compare the topology inferred from a Total Evidence matrix in both Likelihood (RAxML) and Bayesian (MrBayes) and the topology of the Bayesian consensus tree is much better supported than the one of the best ML tree. Specifically, the ML tree has 7/28 nodes have a "fair" bootstrap support ($>$ 75) and 3/28 with a good support ($>$ 90) as the Bayesian consensus tree has 17/25 nodes with a "fair" posterior probability ($>$80) and 14/25 with a good posterior probability ($>$95). We therefore used this reference to strengthen our claim of the superiority of the Bayesian consensus tree over the Maximum Likelihood one.
Secondly, for the question about molecular scaffolds, from the few studies clearly specifying their topological constraint method, only Beck and Lee 2013 (Proceedings B) seems to compare an analysis with and without taxonomic constraint on divergence time and rates estimation (the constraint seems to have no effect) but the branch length aspect is not covered in our study.

See reviewer's comment 11 above for the effective changes in the revised manuscript

% -----------------
% Specific Comment 53
% -----------------
\item{\textcolor{blue}{\textbf{line 628:} use of computer cluster may be better placed in the methods, depending on the norm for MPE.}}

### 2015-08-02 Manuscript/MPESubmission/Revision/TEM_revised_manuscript.tex
Conclusion comment (11)

\section{Conclusions}
Missing data in Total Evidence matrices is not a problem for recovering the ``best'' tree topology as long as enough living and fossil taxa in the matrix have data for overlapping morphological characters.
When missing data increases in any of our missing data parameters ($M_{L}$, $M_{F}$ or $N_{C}$), it reduces support for the placement of fossil taxa and increases the displacement of wildcard taxa.
\hl{Therefore we advise that one should focus on having morphological characters coded for a large amount of living taxa present in the matrix (i.e. 50\%) for accurately combining both living and fossil species in phylogenies.
Doing so grants a good overlap of morphological characters between living and fossil taxa, allowing the fossil taxa to be positioned relatively to the living taxa based on their actual shared derived characters rather than on simply available data.}

>>>>

\section{Conclusions}
\hl{Previous studies have explored the effect of missing morphological data when combining living and fossil species in phylogenies but where limited by their empirical approach or their aspects of missing data }\citep{Wiens01102005,pattinsonphylogeny2014}.
\hl{This study proposes a more in depth approach where missing data is generated on simulated data and along three clear missing-data parameters ($M_{L}$, $M_{F}$ or $N_{C}$).
This allowed use to confirm previous results that missing data can be specially deleterious in small sized matrices }\citep{wiensmissing2003}\hl{ but also relieved a yet unexplored aspect: the crucial importance of coding morphological data for living species.}
\hl{In fact, missing data in Total Evidence matrices is not a problem for recovering the ``best'' tree topology as long as enough living and fossil taxa in the matrix have data for overlapping morphological characters.
When missing data increases in any of our missing data parameters ($M_{L}$, $M_{F}$ or $N_{C}$), it reduces support for the placement of fossil taxa and increases the displacement of wildcard taxa.
Therefore we advise that one should focus on having morphological characters coded for a large amount of living taxa present in the matrix (i.e. 50\%) for accurately combining both living and fossil species in phylogenies.
Doing so grants a good overlap of morphological characters between living and fossil taxa, allowing the fossil taxa to be positioned relatively to the living taxa based on their actual shared derived characters rather than on simply available data.}
